Refunctionalization of Abstract Abstract Machines (Functional Pearl)

### 2018-03-17 paper.tex
update

The abstract syntax in Scala is shown as follows. We assume the source
program conforms ANF's convention, and not force it in the term structure
of Scala constructs.

@@

In this section, we present the version that uses delimited controls.

\todo{Update the code}
\subsection{Un-CPS by Delimited Controls} \label{uncps}

@@

\textbf{Nondeterminism Abstractions, Again.}
The first operator @nd@ is simplified from the previous one. 
Now @nd@ only takes a function @f@ that applies on elements of type @T@, and does recursive call on itself over elements in set @ts@.

\begin{lstlisting}
def nd[T,S](ts: Set[T], acc: Set[S], f: T => Set[S]): Set[S] = {
  if (ts.isEmpty) acc
  else nd(ts.tail, acc ++ vss, f(ts.head))
}
\end{lstlisting}

The second operator @ndcps@ is implemented with delimited controls but simply calls our first operator @nd@.
\begin{lstlisting}
def ndcps[T,S](ts: Set[T], acc: Set[S]): T @cps[Set[S]] = shift { f: (T => Set[S]) =>
  nd(ts, acc, f)
}
\end{lstlisting}

The function @ndcps@ takes two arguments: a set of elements with type @T@, and an initial accumulated value of type @Set[S]@.
The @ndcps@ function iteratively returns a value of type @T@ with annotated by
\verb|@cps[Set[T]]| that denotes the final returned value of @ndcps@ is type @Set[T]@.
The @f@ introduced by the @shift@ operator is the delimited continuation at the call-site of @ndcps@.

We add an annotation \verb|@cps[Ans]| on the return type of the @aeval@ function.
Now, for every nondeterministic choice in the abstract interpreter, we can simply call @ndcps@ on them and write the program sequentially as a concrete interpreter.
The cache is still accumulated by the next call to @ndcps@ or @aeval@, given the fact that the abstract interpreter should always use the latest cache.

\begin{lstlisting}
def aeval(e: Expr, env: Env, store: BStore, time: Time): Ans @cps[Ans] = {
  val new_time = (e::time).take(k)
  e match {
    case Let(x, App(f, ae), e) =>
      val closures = atomicEval(f, env, store).asInstanceOf[Set[Clos]]
      val Clos(Lam(v, body), c_env) = ndcps[Clos, VS](closures, Set[VS]())
      val vbaddr = allocBind(v, new_time)
      val new_cenv = c_env + (v -> vbaddr)
      val new_cstore = store.update(vbaddr, aeval(ae, env, store))
      val bodyvss = aeval(body, new_cenv, new_cstore, new_time)
      val VS(vals, time, vsstore) = ndcps[VS, VS](bodyvss, Set[VS]())
      val xbaddr = allocBind(x, time)
      val new_env = env + (x -> xbaddr)
      val new_store = vsstore.update(xbaddr, vals)
      aeval(e, new_env, new_store, time)

    case ae if isAtomic(ae) =>
      val ds = atomicEval(ae, env, store)
      Set(VS(ds, new_time, store))
  }
}
\end{lstlisting}

@@

\begin{lstlisting}
def analyze(e: Expr) = {
  def iter(cache: Cache): Ans = {
    reset {
      val Ans(vss, anscache) = aval(e, Map[String, BAddr](), Store[BAddr, Storable](Map()), List(), cache)
      val initConfig = Config(e, Map[String, BAddr](), Store[BAddr, Storable](Map()), List())
      val new_cache = anscache.outUpdate(initConfig, vss)
      if (new_cache.out == cache.out) { Ans(vss, new_cache) }
      else { iter(Cache(new_cache.out, new_cache.out)) }
    }
  }
  iter(mtCache)
}
\end{lstlisting}

Now we have arrive the end of this series of transformations. Starting from an abstract abstract machine, 
eventually we obtain an abstract interpreter with pushdown control-flows and written in purely direct-style.

@@

@@

In this functional pearl, we fill the gap between small-step abstract abstract machines
and big-step abstract definitional interpreters by developing a series of syntactical transformations.
Among these transformations, linearization turns worklist to another layer of continuations; 
refunctionalization converts the first-order data types representing continuations to higher-order 
functions; and finally un-CPS with delimited controls transform the abstract interpreter to a sequence of 
expressions.

The broader ...\todo{conc}

>>>>

The abstract syntax in Scala is shown as follows. We assume that the source
program conforms to the ANF convention, and we do not enforce it in the term structure
of Scala constructs.

@@

In this section, we present the second version that uses delimited controls.

\subsection{Un-CPS by Delimited Controls} \label{uncps}

@@

\textbf{Nondeterminism Operators, Again.}
The first operator @nd@ is simplified from the previous one, but specialized with type @Ans@ and @Cache@. 
We may still keep the types to be generic, but the purpose of presentation clearly, exposing @Cache@ explictly
reveals that cache should be monotonically accumulated to the rest of computation.
Now @nd@ only takes a function @f@ that applies on type @(T, Cache)@ where value of type @T@ comes from the set,
and @nd@ does recursive call on itself over elements in the set @ts@.
We also move the accumulation operation (@acc++f(ts.head)@) inside of @nd@.

\begin{lstlisting}
def nd[T](ts: Set[T], acc: Ans, k: ((T, Cache)) => Ans): Ans = {
  if (ts.isEmpty) acc
  else nd(ts.tail, acc ++ k(ts.head, acc.cache), k)
}
\end{lstlisting}

The second operator @ndcps@ is implemented with delimited controls but simply calls our 
first operator @nd@ in body.
\begin{lstlisting}
def ndcps[T](ts: Set[T], acc: Ans): (T, Cache) @cps[Ans] = shift { f: (((T, Cache)) => Ans) =>
  nd(ts, acc, f)
}
\end{lstlisting}
The function @ndcps@ takes two arguments: a set of elements with type @T@, 
and an initial accumulated value of type @Ans@.
The @ndcps@ iteratively returns an element from the set which the element has type @T@ and annotated by
\texttt{\@ cps[Ans]}. This annotation denotes the final returned value of @ndcps@ is type @Ans@.
The @f@ introduced by the @shift@ operator is the delimited continuation from the call-site of @ndcps@.

\textbf{}
We add an annotation \texttt{ cps[Ans]} on the return type of the @aeval@ function.
Now, for every nondeterministic choice in the abstract interpreter, we can simply call 
@ndcps@ on them and write the program sequentially as a concrete interpreter.
@ndcps@ also returns the latest cache to its left-hand side definition.
The cache is still accumulated to the next call to @ndcps@ or @aeval@, 
given the fact that the abstract interpreter should always use the latest cache.

\begin{lstlisting}
def aeval(e: Expr, env: Env, store: BStore, time: Time, cache: Cache): Ans @cps[Ans] = {
  val config = Config(e, env, store, time)
  if (cache.outContains(config)) Ans(cache.outGet(config), cache)
  else {
    val new_time = (e::time).take(k)
    val new_cache = cache.outUpdate(config, cache.inGet(config))
    e match {
      case Let(x, App(f, ae), e) =>
        val closures = atomicEval(f, env, store).asInstanceOf[Set[Clos]]
        val (Clos(Lam(v, body), c_env), clscache) = ndcps[Clos](closures, Ans(Set[VS](), new_cache))
        val vbaddr = allocBind(v, new_time)
        val new_cenv = c_env + (v -> vbaddr)
        val new_cstore = store.update(vbaddr, aeval(ae, env, store))
        /* evaluates the function application App(f, ae) */
        val Ans(bodyvss, bodycache) = aeval(body, new_cenv, new_cstore, new_time, clscache)
        val (VS(vals, time, vsstore), vscache) = ndcps[VS](bodyvss, Ans(Set[VS](), bodycache))
        val baddr = allocBind(x, time)
        val new_env = env + (x -> baddr)
        val new_store = vsstore.update(baddr, vals)
        /* evaluates the body expression from the let */
        val Ans(finval, fincache) = aeval(e, new_env, new_store, time, vscache)
        Ans(finval, fincache.outUpdate(config, finval))

      case ae if isAtomic(ae) =>
        val vs = Set(VS(atomicEval(ae, env, store), new_time, store))
        Ans(vs, cache.outUpdate(config, vs))
    }
  }
}
\end{lstlisting}

@@

\begin{lstlisting}
def analyze(e: Expr) = {
  def iter(cache: Cache): Ans = {
    reset {
      val Ans(vss, anscache) = aeval(e, Map[String, BAddr](), Store[BAddr, Storable](Map()), List(), cache)
      val initConfig = Config(e, Map[String, BAddr](), Store[BAddr, Storable](Map()), List())
      val new_cache = anscache.outUpdate(initConfig, vss)
      if (new_cache.out == cache.out) { Ans(vss, new_cache) }
      else { iter(Cache(new_cache.out, new_cache.out)) }
    }
  }
  iter(mtCache)
}
\end{lstlisting}

Now we have arrive the end of this series of transformations. Starting from an small-step abstract abstract machine, 
eventually we obtain an big-step definitional interpreter with pushdown control-flows and written in direct-style.

@@

Our work establish the pushdown control-flow analysis through refunctionalization and a
proper caching algorithm. The call/return is naturally matched in the stack of 
the meta language.

@@

In this functional pearl, we fill the gap between small-step abstract abstract machines and big-step abstract definitional interpreters by developing a series of syntactical transformations.
Among these transformations, linearization turns a worklist into another layer of continuations; refunctionalization converts the first-order data types representing continuations to higher-order functions; and finally un-CPS with delimited control transforms the abstract interpreter into a sequence of expressions, which looks no different to a concrete interpreter.

We show the correspondence not only exists between concrete semantics artifacts but also exists between abstract semantics artifacts. An interesting open question would be whether there also exists a correspondence of static analyses formalized in different denotational style and operational style.
